Fast, Scalable Phrase-Based SMT Decoding

### 2016-05-03 acl.2016/acl2016.tex
changes

For many users, decoding is the most time-critical part of the translation process. Making use of the multiple cores that are now ubiquitous in todays servers is a common strategy to ameliorate this issue. However, it has been noticed that the Moses decoder, amongst others, is unable to efficiently use multiple cores~\cite{mfernandez2016boosting}. That is, decoding speed does not substantially increase when more cores are used, in fact, it may actually \emph{decrease} when using more cores. There have be speculation on the causes of the inefficiency as well as potential remedies.

@@

There are a number of phrase-based decoding implementations, many of which implements the phrase-based extensions above.

@@

This paper focuses on overcoming the shortcomings of existing decoding implementations on multicore servers.

@@

The standard feature functions in the phrase-based model include:
\begin{enumerate}
  \item \vspace{-2 mm} log transforms translation model probabilities, $p_{TM}(t|s) $ and $p_{TM}(s|t)$, and word-based translation probabilities $p_w(t|s)$ and $p_w(s|t)$,
  \item \vspace{-2 mm} log transforms of the lexicalized re-ordering probabilities,
  \item \vspace{-2 mm} log transforms of the target language model probability $p(t)$, 
  \item \vspace{-2 mm} a distortion penalty
  \item \vspace{-2 mm} a phrase-penalty,
  \item \vspace{-2 mm} a word penalty,
  \item \vspace{-2 mm} an unknown word penalty.
\end{enumerate}
%which we have suffixed with $_{TM} $ to avoid confusion with the overall model probability $ p(t|s) $ and $ p(s|t)$.

Of these feature functions, the last four are simple and do not leave much room for optimization. As have already been mentioned, the language model consumes a large part of decoding time but it has already been heavily optimized in previous works such as ~\newcite{Heafield-kenlm} and ~\newcite{yasuhara-EtAl:2013:EMNLP}.

Therefore, we will focus on optimizing the functionality of the phrase-table and lexicalized re-ordering model.

%\subsection{Language Model}
% TODO
%
\subsection{Translation Model}

The lexicalized reordering model is trained on parallel data, usually requiring random lookups of the model file during decoding. Using the model can increase translation quality at the cost of longer decoding time. Efforts such as~\newcite{junczys_tsd_2012b} have been made to increase the speed of the model.

@@

There are four main optimizations that we shall investigate.

@@

Thirdly, we will look at for optimizating the phrase-table, specifically in relation to the Probing implementation, focusing on two techniques.

One, the default translation rule caching mechanism in Moses saves the most recently used rules. However, this is a complex mechanism which require locking when timestamping datastructures and periodic clearing of old rules in order in order to reduce memory consumption. It actually makes decoding slower, Table~\ref{tab:phrase-table-cache}. 
\begin{table}[h]
\small
\begin{center}
\begin{tabular}{|l|c|c|} \hline
		& No cache	& Caching \\ \hline
Decoding time  	& 2032 		& 2302 (+13\%) \\ \hline
\end{tabular}
\end{center}
\caption{Decoding time (in sec with 32 threads) when using phrase-table cache}
\label{tab:phrase-table-cache}
\end{table}

@@

Lastly, the lexicalized reordering model require random lookups but query key are the source and target phrase of each translation rule. Rather than storing this model separately, we shall investigate integrating it into the translation model.

@@

We trained a phrase-based system using the Moses toolkit with standard settings. The training data consisted of most of the publicly available Arabic-English data from Opus~\cite{tiedemann2012parallel} containing over 69 million parallel sentences, and tuned on a held out set. The phrase-table was then pruned, keeping only the top 100 entries per source phrase, according to $p(t|s)$. All models files were then binarized; the language models were binarized using KenLM~\cite{Heafield-kenlm}, the phrase table using the Probing phrase-table, lexicalized reordering model using the compact datastructure. These binary formats were choosen for their best-in-class multithreaded performance. Table~\ref{tab:model-files} gives details of the resultant sizes of the model files. For testing decoding speed, we used a subset of the training data, Table~\ref{tab:test-sets}. 
\begin{table}[h]
\begin{center}
\begin{tabular}{|l|c|c|} \hline
		& ar-en	& fr-en \\ \hline
Phrase table  	& 17 	& 5.8 \\
Language model 	& 3.1  	& 1.8 \\ 
Lex re. model	& 2.3	& 637MB \\ \hline
\end{tabular}
\end{center}
\caption{Model sizes in GB}
\label{tab:model-files}
\end{table}

@@

In the second optimatization, we disable the compression. This increase the size of the binary files from 17GB to 23GB but the time saved not needing to decompress the data resulted in a 1.5\% decrease in decoding time with 1 thread and nearly 7\% when the CPUs are saturated, Table~\ref{tab:compression}.
\begin{table}[h]
\small
\begin{center}
\begin{tabular}{|r|r|r|} \hline
\# threads	& Compressed pt & Non-compressed pt \\ \hline
1		& 3052	& 3006 (-1.5\%) \\
5		& 756	& 644 (-14.8\%) \\
10		& 372	& 362 (-2.7\%) \\
15		& 284	& 250 (-12.0\%) \\
20		& 244	& 227 (-7.0\%) \\
25		& 218	& 209 (-4.1\%) \\
30		& 206	& 192 (-6.8\%) \\
35		& 203	& 189 (-6.9\%) \\ \hline
\end{tabular}
\end{center}
\caption{Decoding time (in secs with 32 threads) for compressed and non-compressed phrase-tables}
\label{tab:compression}
\end{table}

@@

Figure~\ref{fig:speed} shows decoding speed against the number of threads, measured in words translated per second. There is a constant increase in decoding speed when mrore threads are used, only decreasing slightly after 16 threads when hyperthreads are used. Overall, decoding is 12.5 times faster than single-threaded decoding when all 16 cores (32 hyperthreads) are fully utilized.

>>>>

For many users, decoding is the most time-critical part of the translation process. Making use of the multiple cores that are now ubiquitous in todays servers is a common strategy to ameliorate this issue. However, it has been noticed that the Moses decoder, amongst others, is unable to efficiently use multiple cores~\cite{mfernandez2016boosting}. That is, decoding speed does not substantially increase when more cores are used, in fact, it may actually \emph{decrease} when using more cores. There have been speculation on the causes of the inefficiency as well as potential remedies.

@@

There are a number of phrase-based decoding implementations, many of which implements the extensions to the phrase-based model described above.

@@

This paper not only focuses faster single-threaded decoding but also on overcoming the shortcomings of existing decoding implementations on multicore servers.

@@

The standard feature functions in the phrase-based model include:
\begin{enumerate}
  \item \vspace{-2 mm} a distortion penalty
  \item \vspace{-2 mm} a phrase-penalty,
  \item \vspace{-2 mm} a word penalty,
  \item \vspace{-2 mm} an unknown word penalty.
  \item \vspace{-2 mm} log transforms of the target language model probability $p(t)$, 
  \item \vspace{-2 mm} log transforms translation model probabilities, $p_{TM}(t|s) $ and $p_{TM}(s|t)$, and word-based translation probabilities $p_w(t|s)$ and $p_w(s|t)$,
  \item \vspace{-2 mm} log transforms of the lexicalized re-ordering probabilities,
\end{enumerate}
%which we have suffixed with $_{TM} $ to avoid confusion with the overall model probability $ p(t|s) $ and $ p(s|t)$.

Of these feature functions, the first four are simple and do not leave much room for optimization. As have already been mentioned, the language model consumes a large part of decoding time but it has already been heavily optimized in previous works such as ~\newcite{Heafield-kenlm} and ~\newcite{yasuhara-EtAl:2013:EMNLP}.

Therefore, we will focus on optimizing only the phrase-table and lexicalized reordering model.

@@

Apart from the two feature function optimization detailed in the last section, we will investigate two other optimizations, all described below.

@@

Thirdly, we will look at optimizing the phrase-table.

We will focus on two techniques. One, the default translation rule caching mechanism in Moses saves the most recently used rules. However, this require locking and periodic clearing of old rules in order to reduce memory consumption resulting in \emph{slower} decoding, Table~\ref{tab:phrase-table-cache}. 
\begin{table}[h]
\small
\begin{center}
\begin{tabular}{|l|c|c|} \hline
		& No cache	& Caching \\ \hline
Decoding time  	& 2032 		& 2302 (+13\%) \\ \hline
\end{tabular}
\end{center}
\caption{Decoding time (in sec with 32 threads) when using phrase-table cache}
\label{tab:phrase-table-cache}
\end{table}

@@

Similar to the phrase-table, the lexicalized reordering model is trained on parallel data, a model file is created during training which is then queried during decoding. Inevitably, the querying results in some loss of decoding speed. Efforts such as~\newcite{junczys_tsd_2012b} have been made to increase the speed of the query.

However, the model's query key are the source and target phrase of each translation rule. Rather than storing the lexicalized reordering model separately, we shall integrating it into the translation model, eliminating the need to query a separate file.

@@

We trained a phrase-based system using the Moses toolkit with standard settings. The training data consisted of most of the publicly available Arabic-English data from Opus~\cite{tiedemann2012parallel} containing over 69 million parallel sentences, and tuned on a held out set. The phrase-table was then pruned, keeping only the top 100 entries per source phrase, according to $p(t|s)$. All model files were then binarized; the language models were binarized using KenLM~\cite{Heafield-kenlm}, the phrase table using the Probing phrase-table, lexicalized reordering model using the compact datastructure. These binary formats were chosen for their best-in-class multithreaded performance. Table~\ref{tab:model-files} gives details of the resultant sizes of the model files. For testing decoding speed, we used a subset of the training data, Table~\ref{tab:test-sets}. 
\begin{table}[h]
\begin{center}
\begin{tabular}{|l|c|c|} \hline
		& ar-en	& fr-en \\ \hline
Phrase table  	& 17 	& 5.8 \\
Language model 	& 3.1  	& 1.8 \\ 
Lex re. model	& 2.3	& 637MB \\ \hline
\end{tabular}
\end{center}
\caption{Model sizes in GB}
\label{tab:model-files}
\end{table}

@@

In the second optimization, we disable the compression. This increase the size of the binary files from 17GB to 23GB but the time saved not needing to decompress the data resulted in a 1.5\% decrease in decoding time with 1 thread and nearly 7\% when the CPUs are saturated, Table~\ref{tab:compression}.
\begin{table}[h]
\small
\begin{center}
\begin{tabular}{|r|r|r|} \hline
\# threads	& Compressed pt & Non-compressed pt \\ \hline
1		& 3052	& 3006 (-1.5\%) \\
5		& 756	& 644 (-14.8\%) \\
10		& 372	& 362 (-2.7\%) \\
15		& 284	& 250 (-12.0\%) \\
20		& 244	& 227 (-7.0\%) \\
25		& 218	& 209 (-4.1\%) \\
30		& 206	& 192 (-6.8\%) \\
35		& 203	& 189 (-6.9\%) \\ \hline
\end{tabular}
\end{center}
\caption{Decoding time (in secs with 32 threads) for compressed and non-compressed phrase-tables}
\label{tab:compression}
\end{table}

@@

Figure~\ref{fig:speed} shows decoding speed against the number of threads, measured in words translated per second. There is a constant increase in decoding speed when more threads are used, only decreasing slightly after 16 threads when hyperthreads are used. Overall, decoding is 12.5 times faster than single-threaded decoding when all 16 cores (32 hyperthreads) are fully utilized.
